Natural variation in teosinte at the domestication locus \emph{teosinte branched1} (\emph{tb1})

### 2014-05-20 manuscript.tex
edits to discussion

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Clark2006, Studer2011}.  \citet{Weber2007} have described significant phenotypic associations between genic markers in \emph{tb1} and lateral branch length and female ear length within a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number in our phenotyping experiments, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. In fact, \emph{tb1} is known to be a single gene in a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007}.

\jri{MBH todo} Variation at \emph{tb1} has also been shown to contribute to phenotypes other than tillering \citep{Clark2006}, and a recent study by \citet{StuderDoebley2012} examined the possibility of an allelic series at the \emph{tb1} locus in teosinte. \citet{StuderDoebley2012} introgressed 9 separate teosinte segments (one from \emph{Zea diploperennis}, and four from both \emph{Zea mays} ssp. \emph{mexicana} and \emph{Zea mays} ssp. \emph{parviglumis}) spanning the \emph{tb1} locus into an isogenic maize background and investigated their effects on previously associated phenotypes. They found that plants with teosinte chromosomal segments had greater tillering than their maize isogenic lines, and that different chromosomal segments of \emph{tb1} confer different amounts of tillering, suggesting that there are multiple genetic factors in this region that affect tillering. However, in addition to elucidating variance in tillering among \emph{tb1} teosinte segments, \citet{StuderDoebley2012} found significant variance among W22 control lines, suggesting that there are other genetic factors aside from alleles at the \emph{tb1} locus that affect tillering in maize. \citet{DoebleyStec1991, DoebleyStec1993} first attempted to map QTL controlling many of the phenotypic differences between domesticated maize and teosinte, and demonstrated the existence of numerous QTL that contribute to the differences in branching architecture between the two. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} have been shown to affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number) \citep{Li2012}. It seems plausible that variation in some of these other loci could have affected tillering in our greenhouse population, and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.

In summary, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that  it would have drifted to high frequency in multiple populations and there may be another explanation for the high frequency we observe in some of our populations. The \emph{Hopscotch} does not appear to have a strong effect reducing tillering in teosinte as it does in maize, and  other loci involved in branching architecture  may play also play roles in the regulation of tillering in teosinte. Finally, although we see no clear evidence of recent strong selection, the high frequency of the \emph{Hopscotch} insertion in a number of populations continues to suggest to us that it plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? } \lev{Are they necessary to include? I'd had them in before in a paragraph but had been voted out} \jri{I'd ditch gt1 tru1 ter1 and maybe just cite some people including phyb etc.}

>>>>

Significant effects of the \emph{Hopscotch} insertion on lateral branch length, number of cupules, and tillering index in domesticated maize have been well documented \citep{Studer2011}.  \citet{Weber2007} have described significant phenotypic associations between markers in and around \emph{tb1} and lateral branch length and female ear length within a sample from 74 natural populations of \emph{parviglumis} \citep{Weber2007}; these data did not include markers from the \emph{Hopscotch} region 66kb upstream of \emph{tb1}. Our study is the first to explicitly examine the phenotypic effects of the \emph{Hopscotch} insertion across a wide collection of individuals sampled from natural populations of teosinte. We have found no significant effect of the \emph{Hopscotch} insertion on tillering index or tiller number, a result that is discordant with its clear phenotypic effects in maize. One interpretation of this result would be that the \emph{Hopscotch} controls tillering in maize \citep{Studer2011}, but tillering in teosinte is affected by variation at other loci. In fact, \emph{tb1} is thought to be part of a complex pathway controlling branching, tillering and other phenotypic traits \citep{KebromBrutnell2007, Clark2006}. A recent study by \citet{StuderDoebley2012} examined variation across traits in a three-taxa allelic series at the \emph{tb1} locus. \citet{StuderDoebley2012} introgressed nine unique teosinte \emph{tb1} segments (one from \emph{Zea diploperennis}, and four each from \emph{mexicana} and \emph{parviglumis}) into an inbred maize background and investigated phenotypic effects. Phenotypes were shown to cluster by taxon, indicating \emph{tb1} potentially played a role in the morphological diversification of \emph{Zea}. Additional analysis suggested tillering index was controlled both by \emph{tb1} and loci elsewhere in the genome.  Clues to the identity of these loci may be found in QTL studies that identified loci controlling branching architecture \citep{DoebleyStec1991, DoebleyStec1993}. Many of these loci (\emph{grassy tillers}, \emph{gt1}; \emph{tassel-replaces-upper-ears1}, \emph{tru1}; \emph{terminal ear1}, \emph{ter1}) have been shown to interact with \emph{tb1} \citep{Whipple2011, Li2012},  and both \emph{tru1} and \emph{ter1} affect the same phenotypic traits as \emph{tb1} \citep{DoebleyStecGustus1995}. \emph{tassel-replaces-upper-ears1} (\emph{tru1}), for example, has been shown to act either epistatically or downstream of \emph{tb1}, affecting both branching architecture (decreased apical dominance) and tassel phenotypes (shortened tassel and shank length and reduced tassel number; \citealt{Li2012}). Variation in these additional loci may have affected tillering in our collections and contributed to the lack of correlation we see between \emph{Hopscotch} genotype and tillering.

In summary, our findings demonstrate that the \emph{Hopscotch} allele is more widespread in populations of \emph{parviglumis} and \emph{mexicana} than previously thought. Analysis of linkage using SNPs from across chromosome 1 does not suggest that the \emph{Hopscotch} allele is present in these populations due to recent introgression; however, it seems unlikely that the insertion would have drifted to high frequency in multiple populations. The \emph{Hopscotch} does not appear to reduce tillering in \emph{parviglumis} as it does in maize. Other loci involved in branching architecture may regulate tillering in teosinte. Finally, we find preliminary evidence of selection on the \emph{tb1} locus in \emph{parviglumis}; this coupled with our observation of high frequency of the \emph{Hopscotch} insertion in a number of populations suggests that the locus plays an ecological role in teosinte. In the future, additional experiments will be needed to examine expression levels of \emph{tb1} and additional loci involved in branching architecture (e.g. \emph{gt1}, \emph{tru1}, and \emph{ter1}) in conjunction with a more exhaustive phenotyping and genotyping assay. \jri{why not Phyb and phya? } \lev{Are they necessary to include? I'd had them in before in a paragraph but had been voted out} \jri{I'd ditch gt1 tru1 ter1 and maybe just cite some people including phyb etc.}
